---
title: "QC Report – CNVs Dataset"
output:
  pdf_document: default
  html_document: default
date: "Jacquemont Lab"
params:
  path_dataset: "/lustre09/project/6008022/LAB_WORKSPACE/DATABASE/CNV/SPARK/Legacy_Array_GRCh38"
  path_CNVANNOTATION_repo: "/lustre09/project/6008022/LAB_WORKSPACE/SOFTWARE/Git_pipeline/CNV-Annotation/"
  path_CNVCALLER_repo: "/lustre09/project/6008022/LAB_WORKSPACE/SOFTWARE/Git_pipeline/CNV-Caller/"
---


<!-- 
Rscript -e 'rmarkdown::render("/lustre09/project/6008022/LAB_WORKSPACE/SOFTWARE/bioutils/bin/cnv_utils/cnv_dataset_report.Rmd", 
params=list(path_dataset="/lustre09/project/6008022/LAB_WORKSPACE/DATABASE/CNV/SPARK/SPARK_Array_GRCh37"), 
output_file="/home/flben/links/projects/rrg-jacquese/flben/report_qc_cnv_dataset/data/SPARK_Array_GRCh37.pdf")'
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(arrow)      # Parquet I/O
library(dplyr)      # Data manipulation
library(tidyr)      # Data reshaping
library(stringr)    # String utilities
library(ggplot2)    # Plotting
library(patchwork)  # Plot composition
library(viridis)    # Color palettes
library(scales)     # Axis formatting
library(rlang)
library(readxl)
library(ggh4x)

## ===============================
## Paths and dataset metadata
## ===============================
path_dataset <- params$path_dataset
dataset_name <- basename(path_dataset)

path_CNVANNOTATION_repo <- params$path_CNVANNOTATION_repo
path_CNVCALLER_repo    <- params$path_CNVCALLER_repo


## ===============================
## Load CNV database
## ===============================
cnvDB <- read_parquet(paste(path_dataset,"/cnvDB.parquet",sep=""))

## Basic dataset statistics
sample_number <- n_distinct(cnvDB$SampleID)

## Remove ambiguous CNVs
cnvDB <- cnvDB %>%
  filter(Type != "MIX") %>%
  filter(ProblematicRegions_Overlap < 0.5)


```

# **Dataset:** `r str_replace_all(dataset_name,"_"," ")`  
### **Date:** `r format(Sys.Date(), "%d/%m/%Y")`
### **Number of samples in the dataset:** `r scales::comma(sample_number)`

\vspace{0.5cm}

# 1- CNV length distribution

This section explores the **size distribution of CNVs** across the dataset, stratified by CNV type
(e.g. deletion, duplication).
Notes: We only study CNV with ProblematicRegions_Overlap < 0.5 to focus on high-confidence calls.

## A- Histogram of CNV lengths

```{r cnv-length-dist,echo=FALSE, warning=F, fig.width=7, fig.height=3}

# Define the positions for annotations
annot_positions <- c(1e3, 1e4, 1e5, 1e6, 1e7, 1e8)  # bp

ggplot(cnvDB, aes(x = Length, fill = Type)) +   # map fill to Type
  geom_histogram(binwidth = 0.1, color = "white", position = "dodge") +  # stacked bars by default
  labs(x = "CNV Length (bp)", y = "Number of CNVs") +
  theme_bw() +
  theme(
    text =  element_text(size = 16),
    plot.title = element_text(size = 14),
    strip.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14) 
  ) +
  annotation_logticks(sides = "bp") +
  scale_x_log10(labels = label_number(scale_cut = cut_si("bp")),
                lim = c(500, 1e7),
                breaks = annot_positions) +
  scale_fill_manual(values = c("steelblue", "tomato","orange"))  # manually set two colors

```


## B- Counts per CNV type and size bin


```{r cnv-length-pie,echo=FALSE, warning=F, fig.width=7, fig.height=2}

# Bin CNV lengths
cnvDB <- cnvDB %>%
  mutate(
    LengthBin = cut(
      Length, 
      breaks = c(0, 1e4, 5e4, 1e5, 1e6, Inf),
      labels = c("1-10kb", "10-50kb", "50-100kb", "100kb-1Mb", ">1Mb"),
      include.lowest = TRUE
    )
  )

# Compute total per Type
total_counts <- cnvDB %>%
  group_by(Type) %>%
  summarise(TotalCount = scales::comma(n()), .groups = "drop")

# Join total count back to the CNV table
cnvDB <- cnvDB %>%
  left_join(total_counts, by = "Type") %>%
  mutate(FacetLabel = paste0(Type, " (N= ", TotalCount, ")"))

bin_counts <- cnvDB %>%
  group_by(LengthBin, FacetLabel) %>%
  summarise(Count = n(), .groups = "drop")

# Fix colors using viridis
length_bin_levels <- levels(factor(c("1-10kb", "10-50kb", "50-100kb", "100kb-1Mb", ">1Mb"),levels=c("1-10kb", "10-50kb", "50-100kb", "100kb-1Mb", ">1Mb")))
fixed_colors <- viridis::viridis(length(length_bin_levels), option = "D")
names(fixed_colors) <- length_bin_levels


ggplot(bin_counts, aes(x = "", y = Count, fill = LengthBin)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0) +
  theme_void() +
  labs(
    fill = "CNV length category"
  ) +
  geom_text(data=bin_counts[bin_counts$Count > 10^(max(log10(bin_counts$Count))-2),],aes(label = Count), position = position_stack(vjust = 0.5), size = 4,color="white") +
  scale_fill_manual(values = fixed_colors) +
  facet_wrap(~FacetLabel, scales = "free",ncol=3)+
  theme(
    plot.title = element_text(size = 14),
    strip.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14) 
  )
```

## C- Counts per SampleID and CNV type

```{r cnv-per-sample,echo=FALSE, warning=F, fig.width=5, fig.height=5}
# 1. Compute counts: All CNVs per sample
counts_all <- cnvDB %>%
  group_by(SampleID, Type) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(Category = "All Sizes")

# 2. Compute counts: Large CNVs (> 50kb) per sample
counts_large <- cnvDB %>%
  filter(Length > 50000) %>%
  group_by(SampleID, Type) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(Category = "> 50 kb")

# 3. Merge data for plotting
boxplot_data <- rbind(counts_all, counts_large)

# Ensure category order for faceting
boxplot_data$Category <- factor(boxplot_data$Category, levels = c("All Sizes", "> 50 kb"))

# 4. Generate 4 boxplots with specific zoom scales
ggplot(boxplot_data, aes(x = Type, y = count, fill = Type)) +
  geom_boxplot(outlier.size = 0.5, outlier.alpha = 0.2) +
  
  # Separate by size category
  facet_wrap(~Category, scales = "free_y", ncol = 1) +
  
  # Configure specific zoom scales per facet
  facetted_pos_scales(
    y = list(
      Category == "All Sizes" ~ scale_y_continuous(limits = c(0, 100), oob = scales::squish),
      Category == "> 50 kb"  ~ scale_y_continuous(limits = c(0, 20), oob = scales::squish)
    )
  ) +
  
  theme_bw() +
  scale_fill_manual(values = c("DEL" = "steelblue", "DUP" = "tomato")) +
  
  labs(
    subtitle = "Specific zoom: 0-100 (Global) and 0-20 (> 50 kb)",
    x = "Variant Type",
    y = "Number of CNVs per sample"
  ) +
  theme(
    text = element_text(size = 14),
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "grey95"),
    legend.position = "none"
  )+
  theme(
    plot.subtitle = element_text(size = 10),
    strip.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14) 
  )
```


\newpage

# 2- Analysis of CNV frequency as a function of cumulative LOEUF

This analysis examines the relationship between **CNV frequency** and the **cumulative LOEUF burden** of genes overlapped by each CNV.

For each CNV, `sum_LOEUF` corresponds to the **sum of LOEUF values** of all *canonical transcripts* whose exons are overlapped by the CNV.
CNV frequency is computed **relative to all CNVs of a given type**, including CNVs that do **not overlap any canonical transcript** (these are excluded from LOEUF binning but included in the denominator).

Under purifying selection, we expect CNVs overlapping **LOEUF-intolerant genes** (low cumulative LOEUF) to be **rarer**, resulting in a decreasing CNV frequency with increasing cumulative LOEUF.

\vspace{0.5cm}

```{r cnv-loeuf-frequency, echo=FALSE, warning=F, fig.width=7, fig.height=4}

## -------------------------------
## Bin cumulative LOEUF
## -------------------------------
cnvDB <- cnvDB %>%
  mutate(
    LOEUF_bin = cut(sum_LOEUF, breaks = seq(0, max(sum_LOEUF, na.rm = TRUE), by = 2), include.lowest = TRUE)
  )

## -------------------------------
## Total CNVs per type (denominator)
## -------------------------------
total_per_type <- cnvDB %>%
  group_by(Type) %>%
  summarise(TotalCNV = n(), .groups = "drop")

## -------------------------------
## CNV frequency per LOEUF bin
## -------------------------------
freq_by_LOEUF <- cnvDB[!is.na(cnvDB$LOEUF_bin),] %>%
  group_by(Type, LOEUF_bin) %>%
  summarise(
    CNV_Count = n(),
    Avg_LOEUF = mean(sum_LOEUF, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(total_per_type, by = "Type") %>%
  mutate(Frequency = CNV_Count / TotalCNV)


## -------------------------------
## Plot
## -------------------------------
ggplot(freq_by_LOEUF, aes(x = Avg_LOEUF, y = Frequency, col = Type)) +
  geom_line(stat = "identity", position = "dodge") +
  labs(
    x = "Average cumulative LOEUF per CNV",
    y = "CNV frequency", 
    title = "CNV frequency as a function of cumulative LOEUF") +
  theme_bw() + geom_point() +
  scale_color_manual(values = c("steelblue", "tomato","orange"))+ xlim(0,50)+
  theme(
    text =  element_text(size = 16),
    plot.title = element_text(size = 14),
    strip.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14) 
  )

```

\newpage

# 3- rCNV frequency comparison with UK Biobank–based studies

This analysis compares **recurrent CNV (rCNV) frequencies** observed in the current dataset
to frequencies reported in large external cohorts, primarily **UK Biobank–based studies**.

For each rCNV:

* Frequency is defined as the **percentage of samples carrying at least one event**
* Comparisons are performed **pairwise** between this dataset and published references
* Analyses are stratified by **rare (< 0.06%)** and **more frequent (> 0.06%)** rCNVs

\vspace{0.5cm}

```{r rcnv-frequency-comparison,echo=FALSE, warning=F, fig.width=9, fig.height=6}

## -------------------------------
## Load external rCNV frequencies
## -------------------------------
freq_rCNV <- read_excel(paste(path_CNVANNOTATION_repo,"/resources/rCNV/docs/frequency_comparison.xlsx",sep=""))

group_rCNV <- read.delim(paste(path_CNVANNOTATION_repo,"/resources/rCNV/geneset_per_rCNV.tsv",sep=""))
rownames(group_rCNV) = group_rCNV$rCNV_ID

freq_rCNV = freq_rCNV[,c("Reference","Group_ID","Frequency")]

freq_rCNV = freq_rCNV[freq_rCNV$Reference %in% c("Kendall et al, 2019","Crawford et al, 2019","supp_cecile_2025"),]
freq_rCNV$Reference <- str_replace_all(freq_rCNV$Reference, ",", "")



## -------------------------------
## Compute rCNV frequencies
## -------------------------------
number_samples <- n_distinct(cnvDB$SampleID)

base_id <- str_replace_all(cnvDB$rCNV_ID, "_del|_dup", "")
suffix  <- str_extract(cnvDB$rCNV_ID, "_del|_dup")

cnvDB$Group_ID <- paste0(
  group_rCNV[base_id, "Group_ID"],
  suffix
)


rCNV_stats <- cnvDB %>%
  group_by(Group_ID) %>%
  summarise(
    Reference = dataset_name,
    Count = n(),                     # how many times this rCNV appears
    Frequency = n_distinct(SampleID) / number_samples*100,  # fraction of samples carrying it
    .groups = "drop"
  )

rCNV_stats = rCNV_stats[,c("Reference","Group_ID","Frequency")]

merged_bins = rbind(rCNV_stats,freq_rCNV)


## -------------------------------
## Merge internal and external data
## -------------------------------
merged_bins <- merged_bins %>%
  filter(Frequency != "-" & !is.na(Frequency)) %>%
  mutate(Frequency = as.numeric(Frequency))

merged_bins <- merged_bins %>%
  mutate(
    Reference = gsub("[ ,]", "_", Reference)
  )

freq_wide <- merged_bins %>%
  pivot_wider(names_from = Reference, values_from = Frequency)


## -------------------------------
## Pairwise regression plots
## -------------------------------
plot_list <- list()
x_col = dataset_name
freq_threshold <- 0.06

for(y_col in colnames(freq_wide)[c(-1,-2)]) {
  
  for (label in c("Rare", "Common")) {
    
    plot_data <- freq_wide %>%
      filter(!is.na(.data[[x_col]]) & !is.na(.data[[y_col]])) %>%
      filter(
        if (label == "Rare") .data[[x_col]] <= freq_threshold
        else .data[[x_col]] > freq_threshold
      )
    
    
    lm_fit <- lm(plot_data[[y_col]] ~ plot_data[[x_col]])
    r2 <- summary(lm_fit)$r.squared
    
    p <- ggplot(plot_data, aes(x = .data[[x_col]], y = .data[[y_col]])) +
      geom_abline() +
      geom_point(pch = 21, size = 3, fill = "darkgrey", alpha = 0.8) +
      labs(
        x = paste0(str_replace_all(x_col,"_"," "), " frequency (%)"),
        y = paste0(str_replace_all(y_col,"_"," "), " frequency (%)"),
        title = paste0(
          label, " rCNVs (", 
          ifelse(label == "Rare", "<=", ">"),
          " 0.06%)\nR² = ", round(r2, 3)
        )
      ) +
      theme_bw() +
      theme(
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 11),
        plot.title = element_text(size = 11)
      )
    
    plot_list[[paste(y_col, label)]] <- p
  }
}

wrap_plots(plot_list, ncol = 3)


```

\newpage

# 4- Mendelian precision analysis

This analysis evaluates the **Mendelian precision** of CNV calls using parent–child trios.
For each CNV detected in the child, we assess whether a corresponding event is observed in at least one parent, given an overlap threshold of **25% reciprocal overlap**.

Mendelian precision is defined as the **fraction of child CNVs also observed in a parent**, and is reported as a function of **CNV length** and **CNV type**.
This metric provides a proxy for **call reliability**, with higher precision expected for larger CNVs.

The analysis is performed only if a valid `sampleDB.tsv` file containing trio information
(`SampleID`, `FatherID`, `MotherID`) is available.


\vspace{0.5cm}

```{r mendelian-precision,echo=FALSE, warning=F, fig.width=6, fig.height=4}

sampleDB_path <- file.path(path_dataset, "sampleDB.tsv")

if (!file.exists(sampleDB_path)) {
  message("Mendelian precision analysis skipped: 'sampleDB.tsv' not found.")
} else {
  
  sampleDB <- read.delim(sampleDB_path)
  
  required_cols <- c("SampleID", "FatherID", "MotherID")
  missing_cols <- setdiff(required_cols, colnames(sampleDB))
  
  if (length(missing_cols) > 0) {
    message(
      "Mendelian precision analysis skipped: sampleDB.tsv is missing required columns: "
    )
    message(
      paste(missing_cols, collapse = ", ")
    )
  } else {
    
    ## -------------------------------
    ## Run trio inheritance annotation
    ## -------------------------------
    cmd <- paste(path_CNVCALLER_repo,"/bin/make_report/cnv_trio_inheritance.py",sep="")
    
    system2(
      "python3",
      args = c(
        cmd,
        "-c", paste(path_dataset,"/cnvDB.parquet",sep=""),
        "-p", paste(path_dataset,"/sampleDB.tsv",sep=""),
        "-t", "Type",
        "-f", "0.25"
      )
    )
    
    ## -------------------------------
    ## Load annotated CNVs
    ## -------------------------------
    cnvDB_trio <- read.delim("annotated_child_cnv.tsv") %>%
      filter(Type != "MIX")
    
    number_trio <- n_distinct(cnvDB_trio$SampleID)
    
    ## -------------------------------
    ## Bin CNV lengths
    ## -------------------------------
    cnvDB_trio <- cnvDB_trio %>%
      mutate(
        LengthBin = cut(
          Length, 
          breaks = c(0, 1e4, 5e4, 1e5, 1e6, Inf),  # in bp
          labels = c("1-10kb", "10-50kb", "50-100kb", "100kb-1Mb", ">1Mb"),
          include.lowest = TRUE
        )
      )
    
    
    ## -------------------------------
    ## Mendelian precision per bin
    ## -------------------------------
    mendelian_precision <- cnvDB_trio %>%
      group_by(LengthBin, Type) %>%
      summarise(
        Total_CNVs = n(),
        Avg_Length = mean(Length),
        Observed_in_Parent = sum(Observed_in_Parent_0.25 == "true", na.rm = TRUE),
        Mendelian_Precision = Observed_in_Parent / Total_CNVs,
        SE = sqrt(Mendelian_Precision * (1 - Mendelian_Precision) / Total_CNVs),
        .groups = "drop"
      )
    
    file.remove("annotated_child_cnv.tsv")
    
    ## -------------------------------
    ## Plot
    ## -------------------------------
    ggplot(mendelian_precision, aes(x = LengthBin,y=Mendelian_Precision, color = Type)) +   # map fill to Type
      theme_bw() + geom_errorbar(aes(ymin = Mendelian_Precision - SE, ymax = Mendelian_Precision + SE), width = 0.1) +
      geom_point(size=2) +
      labs(
        title = paste0("Mendelian precision (", scales::comma(number_trio), " trios)"),
        x = "CNV length category",
        y = "Mendelian precision"
      ) + geom_text(aes(label=round(Mendelian_Precision,2)) ,hjust=-0.3) +
      scale_color_manual(values = c("steelblue", "tomato","orange"))  + ylim(0,1)+
      theme(
        text =  element_text(size = 16),
        plot.title = element_text(size = 14),
        axis.text.x = element_text(size = 12, angle=30, vjust=0.5),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14) 
      )
    
    
  }
}
```
