---
title: "QC Report – CNVs Datasets Comparison"
output:
  pdf_document: default
  html_document: default
date: "Jacquemont Lab"
params:
  cnvDB_list: "/lustre09/project/6008022/LAB_WORKSPACE/DATABASE/CNV/SPARK/SPARK_WGS;/lustre09/project/6008022/LAB_WORKSPACE/DATABASE/CNV/SPARK/SPARK_Array_GRCh37;/lustre09/project/6008022/LAB_WORKSPACE/DATABASE/CNV/SPARK/Legacy_Array_GRCh38"
  path_CNVANNOTATION_repo: "/lustre09/project/6008022/LAB_WORKSPACE/SOFTWARE/Git_pipeline/CNV-Annotation/"
  path_CNVCALLER_repo: "/lustre09/project/6008022/LAB_WORKSPACE/SOFTWARE/Git_pipeline/CNV-Caller/"
  path_bioutils_repo: "/lustre09/project/6008022/LAB_WORKSPACE/SOFTWARE/bioutils/"
---
<!-- 
Rscript -e 'rmarkdown::render("/lustre09/project/6008022/LAB_WORKSPACE/SOFTWARE/bioutils/bin/cnv_utils/cnv_datasets_comparison_report.Rmd",
params=list(cnvDB_list="/lustre09/project/6008022/LAB_WORKSPACE/DATABASE/CNV/SPARK/SPARK_WGS;/lustre09/project/6008022/LAB_WORKSPACE/DATABASE/CNV/SPARK/SPARK_Array_GRCh37;/lustre09/project/6008022/LAB_WORKSPACE/DATABASE/CNV/SPARK/Legacy_Array_GRCh38"),
output_file="/lustre09/project/6008022/flben/report_qc_cnv_dataset/data/compare.pdf")'
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(duckdb)
library(arrow)      # Parquet I/O
library(dplyr)      # Data manipulation
library(tidyr)      # Data reshaping
library(stringr)    # String utilities
library(ggplot2)    # Plotting
library(patchwork)  # Plot composition
library(viridis)    # Color palettes
library(scales)     # Axis formatting
library(rlang)
library(readxl)
library(ggh4x)

cnvDB_list = str_split_1(params$cnvDB_list,";")

sample_shared <- Reduce(
  intersect,
  lapply(cnvDB_list, function(x) unique( read_parquet(paste(x, "/cnvDB.parquet",sep=""))$SampleID))
)

sample_number = n_distinct(sample_shared)

```


# **Datasets compared:**   
### `r paste(sapply(cnvDB_list,basename),collapse=" vs ")`  
### **Date:** `r format(Sys.Date(), "%d/%m/%Y")`
### **Number of samples in common between datasets:** `r scales::comma(sample_number)`



```{r analysis, echo=FALSE, warning=F}

## Identify shared sample
length_df = data.frame()
cnv_per_sample_df = data.frame()
loeuf_df = data.frame()
rCNV_df = data.frame()
MP_df = data.frame()
fixed_colors_graph=c()

length_cnvDB = data.frame()

find_sampleDB = sapply(cnvDB_list, function(x) if (file.exists(paste(x,"/sampleDB.tsv",sep=""))){
  if (all(c("SampleID","FatherID","MotherID") %in% colnames(read.delim(paste(x,"/sampleDB.tsv",sep=""))))){
    return(T)
  }
})


for (path_dataset in cnvDB_list){
  dataset_name = str_split_1(path_dataset,"/")[length(str_split_1(path_dataset,"/"))]
  
  # Read the Parquet file
  cnvDB <- read_parquet(paste(path_dataset,"/cnvDB.parquet",sep=""))
  
  # Remove rows with Type == "MIX"
  cnvDB <- cnvDB %>%
    filter(Type != "MIX") %>%
    filter(ProblematicRegions_Overlap < 0.5)
  
  
  cnvDB = cnvDB[cnvDB$SampleID %in% sample_shared, ]
  
  # Bin CNV lengths
  cnvDB <- cnvDB %>%
    mutate(
      LengthBin = cut(
        Length, 
        breaks = c(0, 1e4, 5e4, 1e5, 1e6, Inf),
        labels = c("1-10kb", "10-50kb", "50-100kb", "100kb-1Mb", ">1Mb"),
        include.lowest = TRUE
      )
    )
  
  cnvDB$dataset = dataset_name
  
  length_cnvDB = rbind(length_cnvDB , cnvDB[,c("Length","Type","SampleID","dataset")] )
  
  # Compute total per Type
  total_counts <- cnvDB %>%
    group_by(Type) %>%
    summarise(TotalCount = scales::comma(n()), .groups = "drop")
  
  # Join total count back to the CNV table
  cnvDB <- cnvDB %>%
    left_join(total_counts, by = "Type") %>%
    mutate(FacetLabel = paste0(str_replace_all(dataset,"_"," "), "\n", Type, " (N= ", TotalCount, ")"))
  
  bin_counts <- cnvDB %>%
    group_by(LengthBin, FacetLabel) %>%
    summarise(Count = n(), .groups = "drop")
  
  # Calculate CNVs per sample for each Type
  # 1. Compute counts: All CNVs per sample
  counts_all <- cnvDB %>%
    group_by(SampleID, Type) %>%
    summarise(count = n(), .groups = "drop") %>%
    mutate(Category = "All Sizes")

  # 2. Compute counts: Large CNVs (> 50kb) per sample
  counts_large <- cnvDB %>%
    filter(Length > 50000) %>%
    group_by(SampleID, Type) %>%
    summarise(count = n(), .groups = "drop") %>%
    mutate(Category = "> 50 kb")

  # 3. Merge data for plotting
  boxplot_data <- rbind(counts_all, counts_large)

  # Ensure category order for faceting
  boxplot_data$Category <- factor(boxplot_data$Category, levels = c("All Sizes", "> 50 kb"))
  
  boxplot_data$dataset = dataset_name
  
  cnv_per_sample_df = rbind(cnv_per_sample_df, boxplot_data)

  # Fix colors using viridis
  length_bin_levels <- levels(factor(c("1-10kb", "10-50kb", "50-100kb", "100kb-1Mb", ">1Mb"),levels=c("1-10kb", "10-50kb", "50-100kb", "100kb-1Mb", ">1Mb")))
  fixed_colors <- viridis::viridis(length(length_bin_levels), option = "D")
  names(fixed_colors) <- length_bin_levels
  
  bin_counts$dataset = dataset_name
  
  length_df = rbind(length_df, bin_counts)
  fixed_colors_graph = append(fixed_colors_graph, fixed_colors)
  
  ### LOEUF
  
  cnvDB <- cnvDB %>%
    mutate(
      LOEUF_bin = cut(sum_LOEUF, breaks = seq(0, max(sum_LOEUF, na.rm = TRUE), by = 2), include.lowest = TRUE)
    )
  
  # Total CNVs per Type
  total_per_type <- cnvDB %>%
    group_by(Type) %>%
    summarise(TotalCNV = n(), .groups = "drop")
  
  # Count CNVs per bin and compute average LOEUF per bin
  freq_by_LOEUF <- cnvDB[!is.na(cnvDB$LOEUF_bin),] %>%
    group_by(Type, LOEUF_bin) %>%
    summarise(
      CNV_Count = n(),
      Avg_LOEUF = mean(sum_LOEUF, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    left_join(total_per_type, by = "Type") %>%
    mutate(Frequency = CNV_Count / TotalCNV)
  
  freq_by_LOEUF$dataset = dataset_name
  
  loeuf_df = rbind(loeuf_df, freq_by_LOEUF)
  
  
  ## rCNV
  
  group_rCNV <- read.delim(paste(params$path_CNVANNOTATION_repo,"/resources/rCNV/geneset_per_rCNV.tsv",sep=""))
  rownames(group_rCNV) = group_rCNV$rCNV_ID
  
  number_sample = length(unique(cnvDB$SampleID))
  
  cnvDB_rCNV = cnvDB[!is.na(cnvDB$rCNV_ID),]
  
  base_id <- str_replace_all(cnvDB_rCNV$rCNV_ID, "_del|_dup", "")
  suffix  <- str_extract(cnvDB_rCNV$rCNV_ID, "_del|_dup")
  
  cnvDB_rCNV$Group_ID <- paste0(
    group_rCNV[base_id, "Group_ID"],
    suffix
  )
  
  
  rCNV_stats <- cnvDB_rCNV %>%
    group_by(Group_ID) %>%
    summarise(
      Dataset = dataset_name,
      Count = n(),                     # how many times this rCNV appears
      Frequency = n_distinct(SampleID) / number_sample * 100,  # fraction of samples carrying it
      .groups = "drop"
    )
  
  rCNV_stats = rCNV_stats[,c("Dataset","Group_ID","Frequency")]
  
  rCNV_df = rbind(rCNV_df , rCNV_stats)
  
  
  ### Mendelian Precision

  if (length(unlist(find_sampleDB)) > 0)  {
    sampleDB_path = paste(names(unlist(find_sampleDB))[1],"/sampleDB.tsv",sep="")
    if (file.exists(sampleDB_path)){
      if (all(c("SampleID","FatherID","MotherID") %in% colnames(read.delim(sampleDB_path)))){
        
        cmd <- paste(params$path_CNVCALLER_repo,"/bin/make_report/cnv_trio_inheritance.py",sep="")
        
        # Connect to an in-memory DuckDB database
        con <- dbConnect(duckdb::duckdb(), dbdir = ":memory:")
        
        # Push shared samples into DuckDB
        sample_df <- data.frame(SampleID = sample_shared)
        dbWriteTable(con, "sample_shared", sample_df, temporary = TRUE)
          
        # Filter CNVs to include only shared samples
        dbExecute(con, sprintf("
          CREATE TEMP TABLE cnv_gene_DB AS
          SELECT c.*
          FROM read_parquet('%s/cnvDB.parquet') c
          INNER JOIN sample_shared s
            ON c.SampleID = s.SampleID
          WHERE ProblematicRegions_Overlap < 0.5
        ", path_dataset))
        
        # Save CNV-gene table for this dataset
        dbExecute(con, sprintf("
          COPY cnv_gene_DB
          TO 'cnv_gene_DB_%s.parquet'
          (FORMAT PARQUET)
        ", dataset_name))
        
        # Disconnect from DuckDB
        dbDisconnect(con, shutdown = TRUE)

        system2(
          "python3",
          args = c(
            cmd,
            "-c", paste("cnv_gene_DB_",dataset_name,".parquet",sep=""),
            "-p", sampleDB_path,
            "-t", "Type",
            "-f", "0.25"
          )
        )
        
        unlink(paste("cnv_gene_DB_",dataset_name,".parquet",sep=""), force = TRUE)

        cnvDB_trio = read.delim("annotated_child_cnv.tsv")
        
        length(unique(cnvDB_trio$SampleID))
        
        file.remove("annotated_child_cnv.tsv")
        
        # Bin CNV lengths
        cnvDB_trio <- cnvDB_trio %>%
          filter(Type != "MIX") %>%
          mutate(
            LengthBin = cut(
              Length, 
              breaks = c(0, 1e4, 5e4, 1e5, 1e6, Inf),  # in bp
              labels = c("1-10kb", "10-50kb", "50-100kb", "100kb-1Mb", ">1Mb"),
              include.lowest = TRUE
            )
          )
        
        sampleDB = read.delim(sampleDB_path)
        sampleDB = sampleDB[sampleDB$SampleID %in% sample_shared &
                              sampleDB$FatherID %in% sample_shared &
                              sampleDB$MotherID %in% sample_shared
                            ,]
        
        cnvDB_trio = cnvDB_trio[cnvDB_trio$SampleID %in% sampleDB$SampleID, ]
        
        number_trio <- n_distinct(cnvDB_trio$SampleID)
        
        # Assuming cnvDB already has LengthBin
        mendelian_precision <- cnvDB_trio %>%
          group_by(LengthBin, Type) %>%
          summarise(
            Total_CNVs = n(),
            Avg_Length = mean(Length),
            Observed_in_Parent = sum(Observed_in_Parent_0.25 == "true", na.rm = TRUE),
            Mendelian_Precision = Observed_in_Parent / Total_CNVs,
            SE = sqrt(Mendelian_Precision * (1 - Mendelian_Precision) / Total_CNVs),
            .groups = "drop"
          )
        
        mendelian_precision$dataset = dataset_name
        
        MP_df = rbind(MP_df, mendelian_precision)
      }
    }
  }
}

```


\vspace{0.5cm}

# 1- CNV length distribution

This section explores the **size distribution of CNVs** across the dataset, stratified by CNV type
(e.g. deletion, duplication).

Notes: We only study CNV with ProblematicRegions_Overlap < 0.5 to focus on high-confidence calls.

## A- Histogram of CNV lengths

\vspace{0.5cm}

```{r cnv-length-hist, echo=FALSE, warning=F, fig.width=10, fig.height=5}

# Define the positions for annotations
annot_positions <- c(1e3, 1e5, 1e7)  # bp

ggplot( length_cnvDB , aes(x = Length, fill = Type)) +   # map fill to Type
  geom_histogram(binwidth = 0.2, color = "white", position = "dodge") +  # stacked bars by default
  labs(x = "CNV Length (bp)", y = "Number of CNVs") +
  theme_bw() +
  theme(
    text =  element_text(size = 12),
    plot.title = element_text(size = 14),
    strip.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14) 
  ) + facet_wrap(~str_replace_all(dataset,"_"," ")) +
  annotation_logticks(sides = "bp") +
  scale_x_log10(labels = label_number(scale_cut = cut_si("bp")),
                lim = c(500, 1e7),
                breaks = annot_positions) +
  scale_fill_manual(values = c("steelblue", "tomato","orange"))  # manually set two colors

```


## B- Counts per CNV type and size bin

\vspace{0.5cm}

```{r cnv-length-dist, echo=FALSE, warning=F, fig.width=10, fig.height=10}

ggplot(length_df, aes(x = "", y = Count, fill = LengthBin)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0) +
  theme_void() +
  labs(
    fill = "CNV length category"
  ) +
  geom_text(data=length_df[length_df$Count > 10^(max(log10(length_df$Count))-2),],aes(label = Count), position = position_stack(vjust = 0.5), size = 4,color="white") +
  scale_fill_manual(values = fixed_colors) +
  facet_wrap(~FacetLabel, scales = "free",nrow=length(cnvDB_list))+
  theme(
    legend.position = "top",
    plot.title = element_text(size = 14),
    strip.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14) 
  )

```


## C- Counts per SampleID and CNV type

```{r cnv-per-sample,echo=FALSE, warning=F, fig.width=7, fig.height=7}

ggplot(cnv_per_sample_df, aes(x = dataset , y = count, fill = Type)) +
  geom_boxplot(outlier.size = 0.5, outlier.alpha = 0.2) +
  
  # Separate by size category
  facet_wrap(~Category+Type, scales = "free_y", ncol = 2) +
  
  # Configure specific zoom scales per facet
  facetted_pos_scales(
    y = list(
      Category == "All Sizes" ~ scale_y_continuous(limits = c(0, 100), oob = scales::squish),
      Category == "> 50 kb"  ~ scale_y_continuous(limits = c(0, 20), oob = scales::squish)
    )
  ) +
  
  theme_bw() +
  scale_fill_manual(values = c("DEL" = "steelblue", "DUP" = "tomato")) +
  
  labs(
    subtitle = "Specific zoom: 0-100 (Global) and 0-20 (> 50 kb)",
    x = "Variant Type",
    y = "Number of CNVs per sample"
  ) +
  theme(
    text = element_text(size = 14),
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "grey95"),
    legend.position = "none"
  )+
  theme(
    plot.subtitle = element_text(size = 10),
    strip.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    axis.text.x = element_text(size = 12, angle=30, vjust=0.5),
    legend.title = element_text(size = 14) 
  )
```

\newpage

# 2- Analysis of CNV frequency as a function of cumulative LOEUF

This analysis examines the relationship between **CNV frequency** and the **cumulative LOEUF burden** of genes overlapped by each CNV.

For each CNV, `sum_LOEUF` corresponds to the **sum of LOEUF values** of all *canonical transcripts* whose exons are overlapped by the CNV.
CNV frequency is computed **relative to all CNVs of a given type**, including CNVs that do **not overlap any canonical transcript** (these are excluded from LOEUF binning but included in the denominator).

Under purifying selection, we expect CNVs overlapping **LOEUF-intolerant genes** (low cumulative LOEUF) to be **rarer**, resulting in a decreasing CNV frequency with increasing cumulative LOEUF.

\vspace{0.5cm}

```{r cnv-loeuf-frequency, echo=FALSE, warning=F, fig.width=10, fig.height=5}

## -------------------------------
## Plot
## -------------------------------
ggplot(loeuf_df, aes(x = Avg_LOEUF, y = Frequency, col = Type, linetype=dataset, shape=dataset)) +
  geom_line(stat = "identity", position = "dodge") +
  labs(
    x = "Average cumulative LOEUF per CNV",
    y = "CNV frequency", 
    title = "CNV frequency as a function of cumulative LOEUF") +
  theme_bw() + geom_point() +
  scale_color_manual(values = c("steelblue", "tomato","orange"))+ xlim(0,50)+
  theme(
    text =  element_text(size = 16),
    plot.title = element_text(size = 14),
    strip.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14) 
  ) + facet_wrap(~Type)
```


\newpage

# 3- rCNV frequency comparison between datasets

This analysis compares **recurrent CNV (rCNV) frequencies** observed in the current datasets.

For each rCNV:

* Frequency is defined as the **percentage of samples carrying at least one event**
* Comparisons are performed **pairwise** between this dataset and published references
* Analyses are stratified by **rare (< 0.06%)** and **more frequent (> 0.06%)** rCNVs

\vspace{0.5cm}

```{r rcnv-frequency-comparison,echo=FALSE, warning=F, fig.width=9, fig.height=9}

# ----------------------------------------
# Pairwise rCNV frequency comparison
# ----------------------------------------

# Optional: remove missing or "-" frequencies and convert to numeric
merged_bins <- rCNV_df %>%
  filter(Frequency != "-" & !is.na(Frequency)) %>%
  mutate(Frequency = as.numeric(Frequency))

freq_wide <- merged_bins %>%
  pivot_wider(names_from = Dataset, values_from = Frequency)

freq_threshold <- 0.06
plot_list <- list()

# Dataset columns (exclude Group_ID)
dataset_cols <- setdiff(colnames(freq_wide), "Group_ID")

# All pairwise combinations
pairwise_sets <- combn(dataset_cols, 2, simplify = FALSE)

for (pair in pairwise_sets) {
  
  x_col <- pair[1]
  y_col <- pair[2]
  
  for (label in c("Rare", "Common")) {
    
    plot_data <- freq_wide %>%
      filter(!is.na(.data[[x_col]]) & !is.na(.data[[y_col]])) %>%
      filter(
        if (label == "Rare") .data[[x_col]] <= freq_threshold
        else .data[[x_col]] > freq_threshold
      )
    
    if (length(plot_data[[y_col]]) > 1){
      lm_fit <- lm(plot_data[[y_col]] ~ plot_data[[x_col]])
      r2 <- summary(lm_fit)$r.squared
      
      p <- ggplot(plot_data, aes(x = .data[[x_col]], y = .data[[y_col]])) +
        geom_abline() +
        geom_point(pch = 21, size = 3, fill = "darkgrey", alpha = 0.8) +
        labs(
          x = paste0(str_replace_all(x_col,"_"," "), " frequency (%)"),
          y = paste0(str_replace_all(y_col,"_"," "), " frequency (%)"),
          title = paste0(
            label, " rCNVs (", 
            ifelse(label == "Rare", "<=", ">"),
            " 0.06%)\nR² = ", round(r2, 3)
          )
        ) +
        theme_bw() +
        theme(
          axis.text = element_text(size = 12),
          axis.title = element_text(size = 11),
          plot.title = element_text(size = 11)
        )
      
      plot_list[[paste(y_col, label)]] <- p
    }
  }
}

# ----------------------------------------
# Combine all plots
# ----------------------------------------
wrap_plots(plot_list, ncol = 2)


```


\newpage

# 4- Mendelian precision analysis

This analysis evaluates the **Mendelian precision** of CNV calls using parent–child trios.
For each CNV detected in the child, we assess whether a corresponding event is observed in at least one parent, given an overlap threshold of **25% reciprocal overlap**.

Mendelian precision is defined as the **fraction of child CNVs also observed in a parent**, and is reported as a function of **CNV length** and **CNV type**.
This metric provides a proxy for **call reliability**, with higher precision expected for larger CNVs.

The analysis is performed only if a valid `sampleDB.tsv` file containing trio information
(`SampleID`, `FatherID`, `MotherID`) is available.

\vspace{0.5cm}

```{r mendelian-precision,echo=FALSE, warning=F, fig.width=10, fig.height=5}

if (nrow(MP_df) > 1 ){
  ## -------------------------------
  ## Plot
  ## -------------------------------
  ggplot(MP_df, aes(x = LengthBin,y=Mendelian_Precision, color = Type,
                    shape = dataset,
                    group = dataset)) + geom_line(
                      aes(group = dataset, linetype=dataset),
                      linewidth = 0.8,
                      alpha = 0.7
                    ) +  # map fill to Type
    theme_bw() + geom_errorbar(aes(ymin = Mendelian_Precision - SE, ymax = Mendelian_Precision + SE), width = 0.1) +
    geom_point(size=2) +
    labs(
      title = paste0("Mendelian precision (", scales::comma(number_trio), " trios)"),
      x = "CNV length category",
      y = "Mendelian precision"
    ) + geom_text(aes(label=round(Mendelian_Precision,2)) ,hjust=-0.3) +
    scale_color_manual(values = c("steelblue", "tomato","orange"))  + ylim(0,1)+
    theme(
      text =  element_text(size = 16),
      plot.title = element_text(size = 14),
      axis.text.x = element_text(size = 12, angle=30, vjust=0.5),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14) 
    ) + facet_wrap(~Type) 
  
  
  
} else {
  message("Mendelian precision analysis skipped: 'sampleDB.tsv' not found.")
}


```

\newpage

# 5- Fraction of CNVs reciprocaly found in the other dataset

We compute the fraction of CNVs and their associated genes that are reciprocally observed across different CNV call datasets (using at most 10k samples).

For CNVs, a 25% reciprocal overlap threshold is applied, meaning that only CNVs overlapping by at least 25% in both datasets are considered. For genes, we do not consider the extent of overlap; we only check whether the gene is present or absent in the CNV calls.
To ensure functional relevance and reduce annotation noise, this analysis was restricted to protein-coding biotypes.

\vspace{0.5cm}

```{r overlap, echo=FALSE, warning=F, fig.width=9, fig.height=9}

sample_10k = sample_shared[1:min(10000, length(sample_shared))]

for (path_dataset in cnvDB_list){
  # Extract dataset name from path
  dataset_name = str_split_1(path_dataset,"/")[length(str_split_1(path_dataset,"/"))]
  
  # Connect to an in-memory DuckDB database
  con <- dbConnect(duckdb::duckdb(), dbdir = ":memory:")
  
  # Push shared samples into DuckDB
  sample_df <- data.frame(SampleID = sample_10k)
  dbWriteTable(con, "sample_shared", sample_df, temporary = TRUE)

  # Filter CNVs to include only shared samples
  dbExecute(con, sprintf("
    CREATE TEMP TABLE cnv_DB AS
    SELECT c.*
    FROM read_parquet('%s/cnvDB.parquet') c
    INNER JOIN sample_shared s
      ON c.SampleID = s.SampleID
    WHERE ProblematicRegions_Overlap < 0.5
  ", path_dataset))
  
  # LEFT JOIN CNVs with gene annotations
  dbExecute(con, sprintf("
    CREATE TEMP TABLE cnv_gene_DB AS
    SELECT
      c.*,
      g.*
    FROM cnv_DB c
    LEFT JOIN read_parquet('%s/geneDB.parquet') g
      ON c.CNV_ID = g.CNV_ID
  ", path_dataset))
  
  # Save CNV-gene table for this dataset
  dbExecute(con, sprintf("
    COPY cnv_gene_DB
    TO 'cnv_gene_DB_%s.parquet'
    (FORMAT PARQUET)
  ", dataset_name))
  
  # Disconnect from DuckDB
  dbDisconnect(con, shutdown = TRUE)
}



# Generate all pairwise combinations of datasets
pairwise_sets <- combn(cnvDB_list, 2, simplify = FALSE)

# Initialize a data frame to store overlaps
overlap_db = data.frame()
all_paths <- character()

# Loop over each pair of datasets
for (pair in pairwise_sets) {
  
  db_1 <- pair[1]
  db_2 <- pair[2]
  
  # Construct paths to preprocessed CNV-gene parquet files
  path_a <- file.path(paste0("cnv_gene_DB_", str_split_1(db_1,"/")[length(str_split_1(db_1,"/"))], ".parquet"))
  path_b <- file.path(paste0("cnv_gene_DB_", str_split_1(db_2,"/")[length(str_split_1(db_2,"/"))], ".parquet"))
  
  # Run Python script to compute overlaps
  cmd <- "python3"
  args <- c(paste(params$path_bioutils_repo,"/bin/cnv_utils/cnv_concordance_checker.py",sep=""),
            "--file_a", path_a,
            "--file_b", path_b,
            "--overlap", "0.25")
  
  # Capture output of Python script
  output <- system2(cmd, args, stdout = TRUE, stderr = TRUE)
 
  # Look for lines containing percentages
  pct_lines <- grep("%", output, value = TRUE)
  
  # Extract percentages using regex
  percentages <- str_extract(pct_lines, "\\d+\\.\\d+%")
  
  # Convert to numeric (remove %)
  percentages_num <- as.numeric(str_remove(percentages, "%"))
  
  # Clean dataset names for display
  db_1_name = str_replace_all(str_split_1(db_1,"/")[length(str_split_1(db_1,"/"))],"_"," ")
  db_2_name = str_replace_all(str_split_1(db_2,"/")[length(str_split_1(db_2,"/"))],"_"," ")
  
  # Store results: db_1 vs db_2
  overlap_db = rbind(overlap_db, data.frame(db_1 = db_1_name,
                                            db_2 = db_2_name,
                                            CNV_overlap = percentages_num[2],
                                            GENE_overlap = percentages_num[3]
  ))
  
  # Store reciprocal results: db_2 vs db_1
  overlap_db = rbind(overlap_db, data.frame(db_1 = db_2_name,
                                            db_2 = db_1_name,
                                            CNV_overlap = percentages_num[4],
                                            GENE_overlap = percentages_num[5]
  ))
  # inside loop
  all_paths <- c(all_paths, path_a, path_b)
}

# after loop
unlink(unique(all_paths), force = TRUE)

# Reshape data for plotting with ggplot2
df_long <- overlap_db %>%
  pivot_longer(cols = c(CNV_overlap, GENE_overlap), names_to = "metric", values_to = "value")

# Plot heatmap showing fraction of CNVs/genes in db_1 found in db_2
ggplot(df_long, aes(x = db_2, y = db_1, fill = value)) +  # db_1 in rows, db_2 in columns
  geom_tile(color = "white") +
  facet_wrap(~str_replace_all(metric,"_"," "),ncol=1) +
  scale_fill_gradient(low = "white", high = "darkred", limits = c(0, 100)) +  # percentage scale 0-100
  geom_text(aes(label = value), color = "black", size = 5) +
  theme_minimal() + 
  theme(
    text =  element_text(size = 16),
    plot.title = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle=30, vjust=0.5),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14) ,
    strip.text = element_text(size = 14, face = "bold")  # facet title style
  ) +
  labs(
    x = "Database to check (X)",
    y = "Reference Database (Y)",
    fill = "Overlap %",
    title = "Fraction of CNVs/Genes in Y found in X (%)"
  )
```
